---
title: "Introduction to Pharmacokinetics"
subtitle: "What the Body Does to the Drug"
author: "Based on Boomer.org PHAR 7632 Syllabus"
format:
  revealjs:
    theme: serif
    slide-number: true
    chalkboard: true
    preview-links: auto
    logo: ""
    css: custom.css
    footer: "Introduction to Pharmacokinetics"
    transition: slide
    background-transition: fade
    incremental: false
---

## Learning Objectives {.smaller}

By the end of this lecture, you will be able to:

::: {.incremental}
1. Define pharmacokinetics and explain its clinical importance
2. Describe the four ADME processes
3. Understand and calculate basic PK parameters (Vd, Cl, t½)
4. Interpret a concentration-time curve
5. Distinguish between first-order and zero-order kinetics
6. Explain the concept of bioavailability
:::

::: {.notes}
- Set expectations for what students will learn
- These are foundational concepts for clinical practice
- Reference: Boomer.org Chapter 3
:::

---

# PART 1: Fundamentals

---

## What is Pharmacokinetics? {.smaller}

::: {.columns}
::: {.column width="50%"}
**Definition:**
Study of drug movement through the body over time

**Clinical Importance:**

- Dosing regimen design
- Predicting drug concentrations
- Individualizing therapy
- Understanding drug interactions
:::

::: {.column width="50%"}
```{mermaid}
%%| fig-width: 6
flowchart TD
    A[Drug Administered] --> B[Absorption]
    B --> C[Distribution]
    C --> D[Metabolism]
    D --> E[Excretion]
    C --> E
    style A fill:#e1f5ff
    style E fill:#ffe1e1
```
:::
:::

::: {.notes}
- Contrast with pharmacodynamics (what drug does to body)
- Give example: Why does aspirin work for 4-6 hours?
- Why do some drugs require loading doses?
:::

---

## ADME - The Four Fundamental Processes {.smaller}

::: {.columns}
::: {.column width="60%"}
| Process | Definition | Key Factors |
|---------|------------|-------------|
| **Absorption** | Drug enters bloodstream | Route, formulation, GI pH |
| **Distribution** | Drug disperses through body | Blood flow, Vd, binding |
| **Metabolism** | Drug is chemically transformed | Liver function, CYP enzymes |
| **Excretion** | Drug is removed from body | Renal function, biliary |
:::

::: {.column width="40%"}
**Image Reference:**

- [ResearchGate ADME Diagram](https://www.researchgate.net/figure/Schematic-diagram-showing-the-four-pharmacokinetic-processes-absorption-distribution_fig1_334629263)
- [NCBI ADME Illustration](https://www.ncbi.nlm.nih.gov/books/NBK595006/)
:::
:::

::: {.fragment}
💡 **Interactive:** Which process is fastest for IV vs oral administration?
:::

::: {.notes}
- Each process affects drug concentration over time
- These processes often occur simultaneously
- IV bypasses absorption
:::

---

## The Concentration-Time Relationship {.smaller}

::: {.columns}
::: {.column width="50%"}
**Key Concept:**

- Drug effect relates to concentration at site of action
- We measure plasma concentration as surrogate
- Concentration changes over time due to ADME

**Important Parameters:**

- **Cmax:** Peak concentration
- **Tmax:** Time to peak
- **AUC:** Area under curve (total exposure)
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 4

library(ggplot2)

# Generate data for oral administration
time <- seq(0, 24, by = 0.1)
conc_oral <- 10 * time * exp(-0.3 * time)

# Create data frame
df <- data.frame(time = time, concentration = conc_oral)

# Plot
ggplot(df, aes(x = time, y = concentration)) +
  geom_line(color = "blue", size = 1.2) +
  geom_vline(xintercept = 1/0.3, linetype = "dashed", color = "red") +
  annotate("text", x = 4, y = 3.5, label = "Tmax", color = "red") +
  annotate("text", x = 4, y = 3, label = "Cmax", color = "red") +
  geom_segment(aes(x = 3.33, y = 0, xend = 3.33, yend = 3.68),
               linetype = "dashed", color = "red") +
  labs(title = "Typical Oral Drug Concentration-Time Curve",
       x = "Time (hours)",
       y = "Concentration (mg/L)") +
  theme_minimal() +
  theme(text = element_text(size = 14))
```
:::
:::

**Image References:**

- [Plasma Concentration-Time Curves (ScienceDirect)](https://www.sciencedirect.com/topics/biochemistry-genetics-and-molecular-biology/plasma-concentration-time-curve)
- [ResearchGate Oral Administration Curve](https://www.researchgate.net/figure/Plasma-level-time-curve-for-a-drug-administered-by-oral-route_fig2_295703806)

---

# PART 2: Compartment Models & IV Bolus

---

## Introduction to Compartment Models {.smaller}

::: {.columns}
::: {.column width="50%"}
**Compartment:** Theoretical space where drug distributes uniformly

**One-Compartment Model:**

- Simplest case
- Assumes instantaneous distribution
- Drug behaves as if in single homogeneous space
- Mathematical model, not anatomical
:::

::: {.column width="50%"}
```{mermaid}
%%| fig-width: 6
flowchart LR
    A[Dose] --> B[Central Compartment<br/>Volume = Vd]
    B --> C[Elimination<br/>Rate = k × C]
    style B fill:#fff4e6
    style C fill:#ffe1e1
```
:::
:::

**Useful for:**

- Drugs with rapid distribution
- Initial understanding of PK principles
- Examples: Aminoglycosides

**Resources:**

- [Boomer.org Chapter 4](https://www.boomer.org/c/p4/c04/c04.pdf)
- [Deranged Physiology - Compartment Models](https://derangedphysiology.com/main/cicm-primary-exam/pharmacokinetics/Chapter-201/single-and-multiple-compartment-models-drug-distribution)

---

## IV Bolus - Simplest Administration {.smaller}

::: {.columns}
::: {.column width="50%"}
**IV Bolus:**

- Entire dose given instantly into bloodstream
- Absorption = 100% (bypasses absorption)
- Allows focus on distribution and elimination

**First-Order Elimination Equation:**

$$C(t) = C_0 \cdot e^{-k \cdot t}$$

Where:

- $C(t)$ = concentration at time t
- $C_0$ = initial concentration
- $k$ = elimination rate constant
- $t$ = time
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 5

library(ggplot2)

# Generate data for IV bolus
time <- seq(0, 24, by = 0.1)
C0 <- 100
k <- 0.139
conc_iv <- C0 * exp(-k * time)

df <- data.frame(time = time, concentration = conc_iv)

ggplot(df, aes(x = time, y = concentration)) +
  geom_line(color = "darkblue", size = 1.2) +
  geom_hline(yintercept = C0/2, linetype = "dashed", color = "red", alpha = 0.5) +
  geom_hline(yintercept = C0/4, linetype = "dashed", color = "red", alpha = 0.5) +
  annotate("text", x = 20, y = 55, label = "50% (1 t½)", color = "red") +
  annotate("text", x = 20, y = 28, label = "25% (2 t½)", color = "red") +
  labs(title = "IV Bolus: Exponential Decay (First-Order)",
       x = "Time (hours)",
       y = "Concentration (mg/L)") +
  theme_minimal() +
  theme(text = element_text(size = 14))
```
:::
:::

::: {.notes}
- This exponential equation describes first-order elimination
- Most drugs follow first-order kinetics
:::

---

## KEY PARAMETER 1: Volume of Distribution (Vd) {.smaller}

**Definition:** Theoretical volume needed to account for all drug in the body at the same concentration as plasma

$$V_d = \frac{Dose}{C_0}$$

::: {.columns}
::: {.column width="50%"}
**Clinical Interpretation:**

| Vd Range | Interpretation | Example |
|----------|----------------|---------|
| 5-15 L | Confined to plasma | Warfarin (~8 L) |
| 15-50 L | Extracellular fluid | Many antibiotics |
| >50 L | Extensive tissue distribution | Digoxin (~500 L) |

💡 **Vd > total body water** means extensive tissue binding
:::

::: {.column width="50%"}
```{mermaid}
%%| fig-width: 6
graph TD
    A[Same Dose<br/>100 mg] --> B[Low Vd = 10 L<br/>C₀ = 10 mg/L]
    A --> C[High Vd = 100 L<br/>C₀ = 1 mg/L]
    B --> D[Drug in plasma]
    C --> E[Drug in tissues]
    style B fill:#ffcccc
    style C fill:#ccffcc
```
:::
:::

::: {.fragment}
**Practice:** If 1000 mg gives C₀ = 20 mg/L, what is Vd?

::: {.fragment}
**Answer:** Vd = 1000 mg ÷ 20 mg/L = **50 L**
:::
:::

---

## KEY PARAMETER 2: Clearance (Cl) {.smaller}

**Definition:** Volume of plasma completely cleared of drug per unit time

::: {.columns}
::: {.column width="50%"}
**Formulas:**

$$Cl = k \cdot V_d$$

$$Cl = \frac{Dose}{AUC}$$

**Units:** mL/min or L/h

**Components:**

$$Cl_{total} = Cl_{hepatic} + Cl_{renal} + Cl_{other}$$
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 4

library(ggplot2)

# Clearance concept
organs <- c("Liver", "Kidneys", "Other")
clearance <- c(40, 55, 5)

df <- data.frame(organs = factor(organs, levels = organs),
                 clearance = clearance)

ggplot(df, aes(x = organs, y = clearance, fill = organs)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("#8B4513", "#DC143C", "#4682B4")) +
  labs(title = "Typical Drug Clearance Distribution",
       x = "Eliminating Organ",
       y = "Clearance (mL/min)") +
  theme_minimal() +
  theme(text = element_text(size = 14), legend.position = "none")
```
:::
:::

**Clinical Significance:**

- Affected by **liver function** (hepatic Cl) and **kidney function** (renal Cl)
- Determines **maintenance dose**
- Renal impairment reduces Cl → need dose adjustment

---

## KEY PARAMETER 3: Half-Life (t½) {.smaller}

**Definition:** Time required for plasma concentration to decrease by 50%

::: {.columns}
::: {.column width="50%"}
**Formulas:**

$$t_{1/2} = \frac{0.693}{k}$$

$$t_{1/2} = \frac{0.693 \cdot V_d}{Cl}$$

**Clinical Significance:**

- Determines **dosing interval**
- Time to steady-state ≈ **5 × t½**
- Time to eliminate ≈ **5 × t½**
:::

::: {.column width="50%"}
| Time | Fraction Remaining | % Remaining |
|------|-------------------|-------------|
| 1 t½ | 1/2 | 50% |
| 2 t½ | 1/4 | 25% |
| 3 t½ | 1/8 | 12.5% |
| 4 t½ | 1/16 | 6.25% |
| 5 t½ | 1/32 | 3.1% |
:::
:::

::: {.fragment}
**Practice:** If C₀ = 100 mg/L and t½ = 4 hours:

- C at 4h? ::: {.fragment}**50 mg/L**:::
- C at 8h? ::: {.fragment}**25 mg/L**:::
- C at 12h? ::: {.fragment}**12.5 mg/L**:::
:::

::: {.notes}
- After 5 half-lives, ~97% of drug is eliminated
- Reference: Boomer.org Chapter 5
:::

---

## Relating Vd, Cl, k, and t½ {.smaller}

**The Four Parameters are Interconnected:**

```{mermaid}
%%| fig-width: 8
graph TD
    A[Vd<br/>Volume of Distribution] -.-> C[k<br/>Elimination Rate Constant]
    B[Cl<br/>Clearance] -.-> C
    C -.-> D[t½<br/>Half-Life]
    A -.-> D
    B -.-> D

    E["k = Cl / Vd"] -.-> C
    F["t½ = 0.693 / k"] -.-> D
    G["t½ = (0.693 × Vd) / Cl"] -.-> D

    style A fill:#ffe6e6
    style B fill:#e6f3ff
    style C fill:#fff4e6
    style D fill:#e6ffe6
```

::: {.callout-tip}
## Key Relationships
- **k** depends on both Cl and Vd: $k = Cl / V_d$
- **t½** depends on both Cl and Vd: $t_{1/2} = \frac{0.693 \cdot V_d}{Cl}$
- If Cl normal but t½ long → **Vd must be large**
- If Vd normal but t½ short → **Cl must be high**
:::

---

## First-Order vs Zero-Order Kinetics {.smaller}

::: {.columns}
::: {.column width="50%"}
**First-Order (Most Common):**

- Constant **fraction** eliminated per unit time
- Exponential decay
- t½ is **constant**
- Rate = k × C

**Examples:** Most drugs at therapeutic doses

**Zero-Order (Rare):**

- Constant **amount** eliminated per unit time
- Linear decay
- t½ **changes** with concentration
- Rate = constant

**Examples:** Ethanol, phenytoin, aspirin (high doses)
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 5

library(ggplot2)

time <- seq(0, 10, by = 0.1)

# First-order
first_order <- 100 * exp(-0.3 * time)

# Zero-order
zero_order <- pmax(100 - 10 * time, 0)

df <- data.frame(
  time = rep(time, 2),
  concentration = c(first_order, zero_order),
  kinetics = rep(c("First-Order (Exponential)",
                   "Zero-Order (Linear)"), each = length(time))
)

ggplot(df, aes(x = time, y = concentration,
               color = kinetics, linetype = kinetics)) +
  geom_line(size = 1.2) +
  scale_color_manual(values = c("blue", "red")) +
  labs(title = "First-Order vs Zero-Order Kinetics",
       x = "Time (hours)",
       y = "Concentration (mg/L)",
       color = "Kinetics Type",
       linetype = "Kinetics Type") +
  theme_minimal() +
  theme(text = element_text(size = 12),
        legend.position = "bottom")
```
:::
:::

**Resources:**

- [NCBI: First and Zero Order Kinetics](https://www.ncbi.nlm.nih.gov/books/NBK499866/)
- [Deranged Physiology Comparison](https://derangedphysiology.com/main/cicm-primary-exam/pharmacokinetics/Chapter-337/first-order-zero-order-and-non-linear-elimination-kinetics)

---

# PART 3: IV Infusion & Steady State

---

## Continuous IV Infusion {.smaller}

::: {.columns}
::: {.column width="50%"}
**Concept:**

- Drug administered at constant rate (R₀)
- Builds up to steady-state
- **Input rate = Elimination rate** at steady state

**Formula:**

$$C_{ss} = \frac{R_0}{Cl}$$

**Time to Steady-State:**

- 50% of Css at 1 × t½
- 75% of Css at 2 × t½
- 87.5% of Css at 3 × t½
- **97% of Css at 5 × t½**
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 5

library(ggplot2)

time <- seq(0, 50, by = 0.5)
t_half <- 10
k <- 0.693 / t_half
Css <- 15

# Concentration during infusion
conc <- Css * (1 - exp(-k * time))

df <- data.frame(time = time, concentration = conc)

ggplot(df, aes(x = time, y = concentration)) +
  geom_line(color = "darkgreen", size = 1.2) +
  geom_hline(yintercept = Css, linetype = "dashed",
             color = "red", size = 1) +
  geom_hline(yintercept = Css * 0.5, linetype = "dotted",
             color = "blue", alpha = 0.5) +
  annotate("text", x = 45, y = Css + 0.5,
           label = "Steady-State", color = "red") +
  annotate("segment", x = t_half, xend = t_half,
           y = 0, yend = Css * 0.5,
           linetype = "dashed", color = "blue", alpha = 0.5) +
  annotate("text", x = t_half, y = -1,
           label = "1 t½", color = "blue") +
  annotate("segment", x = 5*t_half, xend = 5*t_half,
           y = 0, yend = Css * 0.97,
           linetype = "dashed", color = "orange", alpha = 0.5) +
  annotate("text", x = 5*t_half, y = -1,
           label = "5 t½", color = "orange") +
  labs(title = "IV Infusion: Rise to Steady-State",
       x = "Time (hours)",
       y = "Concentration (mg/L)") +
  theme_minimal() +
  theme(text = element_text(size = 14))
```
:::
:::

**Used for:** Drugs needing constant levels (dopamine, heparin, propofol)

**Image Reference:** [ResearchGate IV Infusion Curve](https://www.researchgate.net/figure/Plasma-concentration-time-curve-after-IV-Infusion_fig7_295703806)

---

## Loading Dose Concept {.smaller}

::: {.columns}
::: {.column width="50%"}
**Problem:** Takes 5 × t½ to reach steady-state. What if we need therapeutic levels **immediately**?

**Solution:** Loading Dose

$$Loading\ Dose = V_d \times C_{target}$$

**Example:**

- Drug with t½ = 24 hours
- Would take **5 days** to steady-state!
- Loading dose achieves therapeutic level immediately
- Maintenance dose maintains it
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 5

library(ggplot2)

time <- seq(0, 120, by = 1)
t_half <- 24
k <- 0.693 / t_half
Css <- 12

# Without loading dose
no_load <- Css * (1 - exp(-k * time))

# With loading dose
with_load <- ifelse(time == 0, Css,
                    Css * (1 - exp(-k * (time - 1))) +
                    Css * exp(-k * (time - 1)))
with_load[1] <- Css

df <- data.frame(
  time = rep(time, 2),
  concentration = c(no_load, with_load),
  group = rep(c("Without Loading Dose", "With Loading Dose"),
              each = length(time))
)

ggplot(df, aes(x = time, y = concentration,
               color = group, linetype = group)) +
  geom_line(size = 1.2) +
  geom_hline(yintercept = Css * 0.9, linetype = "dashed",
             color = "gray", alpha = 0.5) +
  scale_color_manual(values = c("red", "darkgreen")) +
  labs(title = "Loading Dose: Immediate Therapeutic Level",
       x = "Time (hours)",
       y = "Concentration (mg/L)",
       color = "",
       linetype = "") +
  theme_minimal() +
  theme(text = element_text(size = 12),
        legend.position = "bottom")
```
:::
:::

::: {.callout-warning}
**Risk:** Toxicity if Vd estimated incorrectly!
:::

**Common uses:** Digoxin, phenytoin, some antibiotics

---

# PART 4: Oral Administration & Bioavailability

---

## Oral Administration - A Different Story {.smaller}

::: {.columns}
::: {.column width="50%"}
**Differences from IV:**

- Must be **absorbed** (absorption phase)
- Cmax < C₀ for same dose
- Tmax - time to reach peak
- **First-pass metabolism** reduces amount reaching circulation

**ADME for Oral Drugs:**

- ✅ **A**bsorption - CRITICAL
- ✅ **D**istribution - same as IV
- ✅ **M**etabolism - BEFORE systemic circulation (first-pass)
- ✅ **E**xcretion - same as IV
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 5

library(ggplot2)

time <- seq(0, 24, by = 0.1)

# IV bolus
C0_iv <- 10
k <- 0.2
conc_iv <- C0_iv * exp(-k * time)

# Oral (same dose but F = 0.5)
ka <- 0.6  # absorption rate
conc_oral <- C0_iv * 0.5 * (ka/(ka-k)) *
             (exp(-k*time) - exp(-ka*time))

df <- data.frame(
  time = rep(time, 2),
  concentration = c(conc_iv, conc_oral),
  route = rep(c("IV Bolus", "Oral"), each = length(time))
)

ggplot(df, aes(x = time, y = concentration,
               color = route, linetype = route)) +
  geom_line(size = 1.2) +
  scale_color_manual(values = c("blue", "red")) +
  annotate("text", x = 3, y = 4, label = "Cmax", color = "red") +
  annotate("text", x = 0.5, y = 9, label = "C₀", color = "blue") +
  labs(title = "IV vs Oral: Same Dose, Different Curves",
       x = "Time (hours)",
       y = "Concentration (mg/L)",
       color = "Route",
       linetype = "Route") +
  theme_minimal() +
  theme(text = element_text(size = 14),
        legend.position = "bottom")
```
:::
:::

::: {.notes}
- Oral is most convenient route
- But more complex pharmacokinetics
- Reference: Boomer.org Chapters 7-8
:::

---

## Bioavailability (F) {.smaller}

**Definition:** Fraction of administered dose that reaches systemic circulation unchanged

$$F = \frac{AUC_{oral} \times Dose_{IV}}{AUC_{IV} \times Dose_{oral}}$$

::: {.columns}
::: {.column width="50%"}
**Range:** 0 to 1 (or 0% to 100%)

**Factors Affecting F:**

- First-pass hepatic metabolism ⚠️
- Incomplete absorption
- Drug degradation in GI tract
- Efflux transporters (P-glycoprotein)

**Clinical Examples:**

| Drug | Oral F | Reason |
|------|--------|--------|
| IV drugs | 100% | Direct to circulation |
| Morphine | ~25% | Extensive first-pass |
| Nitroglycerin | <1% | Almost complete first-pass |
| Amoxicillin | ~80% | Good absorption |
:::

::: {.column width="50%"}
```{mermaid}
%%| fig-width: 6
flowchart TD
    A[Oral Dose<br/>100%] --> B[GI Tract]
    B --> C{Absorption}
    C -->|80%| D[Portal Vein]
    C -->|20% Lost| E[Not Absorbed]
    D --> F{Liver<br/>First-Pass}
    F -->|50%| G[Systemic Circulation<br/>F = 40%]
    F -->|50% Lost| H[Metabolized]

    style A fill:#e1f5ff
    style G fill:#e1ffe1
    style E fill:#ffe1e1
    style H fill:#ffe1e1
```
:::
:::

**Dose Adjustment:**

$$Dose_{oral} = \frac{Dose_{IV}}{F}$$

**Resources:**

- [NCBI First-Pass Effect](https://www.ncbi.nlm.nih.gov/books/NBK551679/)
- [ResearchGate First-Pass Diagram](https://www.researchgate.net/figure/The-concept-of-first-pass-metabolism-and-bioavailability-modified-from-Rowland-Tozer_fig3_33935609)

---

## Bioequivalence {.smaller}

**Definition:** Two formulations are bioequivalent if they deliver same amount of drug to circulation at same rate

::: {.columns}
::: {.column width="50%"}
**Criteria:**

- Same **AUC** (extent of absorption)
- Same **Cmax** and **Tmax** (rate of absorption)

**Clinical Importance:**

- Generic vs brand-name drugs
- Changing formulations
- Ensuring therapeutic equivalence

**FDA Requirements:**

- 90% confidence interval
- 80-125% ratio for AUC and Cmax
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 4

library(ggplot2)

time <- seq(0, 24, by = 0.1)
ka <- 0.5
k <- 0.2

# Brand drug
conc_brand <- 10 * (ka/(ka-k)) * (exp(-k*time) - exp(-ka*time))

# Generic (bioequivalent - slightly different curve but same AUC)
ka2 <- 0.55
conc_generic <- 10 * (ka2/(ka2-k)) * (exp(-k*time) - exp(-ka2*time))

df <- data.frame(
  time = rep(time, 2),
  concentration = c(conc_brand, conc_generic),
  formulation = rep(c("Brand", "Generic (Bioequivalent)"),
                    each = length(time))
)

ggplot(df, aes(x = time, y = concentration,
               color = formulation, linetype = formulation)) +
  geom_line(size = 1.2) +
  scale_color_manual(values = c("blue", "red")) +
  labs(title = "Bioequivalent Formulations",
       subtitle = "Similar AUC, Cmax, and Tmax",
       x = "Time (hours)",
       y = "Concentration (mg/L)",
       color = "",
       linetype = "") +
  theme_minimal() +
  theme(text = element_text(size = 12),
        legend.position = "bottom")
```
:::
:::

---

# PART 5: Clinical Applications

---

## Clinical Case: Gentamicin Dosing {.smaller}

::: {.panel-tabset}

### Case Presentation

**Patient:** 70 kg adult with pneumonia

**Drug:** Gentamicin (antibiotic)

**Parameters:**

- Vd = 0.25 L/kg
- t½ = 2.3 hours
- Target Css = 5 mg/L

**Questions:**

1. What infusion rate for steady-state?
2. What loading dose if needed?
3. Time to steady-state without loading dose?

### Calculations

**Step 1: Calculate total Vd**
$$V_d = 0.25\ L/kg \times 70\ kg = 17.5\ L$$

**Step 2: Calculate clearance**
$$k = \frac{0.693}{2.3\ h} = 0.301\ h^{-1}$$
$$Cl = k \times V_d = 0.301 \times 17.5 = 5.27\ L/h$$

**Step 3: Infusion rate**
$$R_0 = C_{ss} \times Cl = 5\ mg/L \times 5.27\ L/h = 26.4\ mg/h$$

### Results

**Answers:**

1. **Infusion rate:** 26.4 mg/h (≈ 26-27 mg/h)

2. **Loading dose:**
   $$LD = V_d \times C_{target} = 17.5\ L \times 5\ mg/L = 87.5\ mg$$

3. **Time to steady-state:**
   $$5 \times t_{1/2} = 5 \times 2.3\ h = 11.5\ hours$$

::: {.callout-tip}
Loading dose saves ~11.5 hours!
:::

:::

---

## Special Populations {.smaller}

::: {.panel-tabset}

### Renal Impairment

**Changes:**

- ↓ **Clearance** (for renally eliminated drugs)
- ↑ **Half-life**
- Need **dose reduction**

**Dose Adjustment:**

$$Dose_{new} = Dose_{normal} \times \frac{CrCl_{patient}}{CrCl_{normal}}$$

**Examples:**

- Aminoglycosides (gentamicin, tobramycin)
- Vancomycin
- Many beta-lactams
- Gabapentin

### Hepatic Impairment

**Changes:**

- ↓ **Metabolism**
- ↓ **Clearance**
- ↑ **Half-life**

**Approach:**

- No simple formula (use Child-Pugh score)
- Generally reduce dose by 25-50%
- Monitor closely

**Examples:**

- Warfarin
- Most benzodiazepines
- Opioids
- Propranolol

### Elderly

**Changes with Aging:**

- ↓ Renal function (even with normal SCr)
- ↓ Lean body mass
- ↑ Body fat percentage
- ↓ Hepatic blood flow
- ↓ Albumin (affects protein binding)

**Approach:**

- **"Start low, go slow"**
- Use lower end of dosing range
- Monitor for adverse effects
- Calculate CrCl (don't rely on SCr alone)

### Obesity

**Dosing Weight Considerations:**

| Drug Type | Weight to Use |
|-----------|---------------|
| Hydrophilic (small Vd) | Ideal Body Weight (IBW) |
| Lipophilic (large Vd) | Total Body Weight (TBW) |
| Compromise | Adjusted Body Weight (ABW) |

**Formulas:**

$$IBW_{male} = 50\ kg + 2.3\ kg\ per\ inch\ over\ 5\ feet$$
$$IBW_{female} = 45.5\ kg + 2.3\ kg\ per\ inch\ over\ 5\ feet$$
$$ABW = IBW + 0.4 \times (TBW - IBW)$$

:::

---

# Summary & Key Concepts

---

## Key Concepts Summary {.smaller}

::: {.columns}
::: {.column width="50%"}
**What We Covered:**

✅ Pharmacokinetics = what body does to drug

✅ ADME processes

✅ Three critical parameters:

- **Vd** - Volume of distribution
- **Cl** - Clearance
- **t½** - Half-life

✅ First-order kinetics (exponential decay)

✅ IV infusion and steady-state

✅ Bioavailability for oral drugs
:::

::: {.column width="50%"}
**The Foundation:**

These basic principles underlie ALL pharmacokinetic calculations and clinical applications

```{mermaid}
%%| fig-width: 6
graph TD
    A[PK Fundamentals] --> B[Dosing Calculations]
    A --> C[TDM]
    A --> D[Drug Interactions]
    A --> E[Special Populations]
    B --> F[Optimal Patient Care]
    C --> F
    D --> F
    E --> F
    style A fill:#ffe6e6
    style F fill:#e6ffe6
```
:::
:::

---

## Essential Formulas - Quick Reference {.smaller}

::: {.columns}
::: {.column width="50%"}
**Basic Parameters:**

$$V_d = \frac{Dose}{C_0}$$

$$Cl = k \times V_d = \frac{Dose}{AUC}$$

$$k = \frac{Cl}{V_d} = \frac{0.693}{t_{1/2}}$$

$$t_{1/2} = \frac{0.693}{k} = \frac{0.693 \times V_d}{Cl}$$
:::

::: {.column width="50%"}
**Clinical Applications:**

$$C(t) = C_0 \times e^{-k \times t}$$

$$C_{ss} = \frac{R_0}{Cl}$$

$$Loading\ Dose = V_d \times C_{target}$$

$$F = \frac{AUC_{oral} \times Dose_{IV}}{AUC_{IV} \times Dose_{oral}}$$
:::
:::

::: {.callout-note}
## Remember
- Always include **units** in calculations
- **5 half-lives** rule for steady-state and elimination
- **Clearance** determines maintenance dose
- **Vd** determines loading dose
:::

---

## What's Next? Advanced Topics {.smaller}

**Build on this foundation:**

::: {.incremental}
1. **Multiple compartment models** (2-compartment, 3-compartment)
2. **Multiple dosing and accumulation** (dosing intervals, fluctuation)
3. **Non-linear kinetics** (Michaelis-Menten, saturable elimination)
4. **Therapeutic drug monitoring** (peak/trough, target levels)
5. **Population pharmacokinetics** (variability, covariates)
6. **Pharmacokinetic/pharmacodynamic modeling** (linking PK to effects)
:::

---

## Practice Problems {.smaller}

::: {.panel-tabset}

### Problem 1

**Question:**

A 1000 mg IV bolus dose yields C₀ = 25 mg/L. The half-life is 6 hours.

Calculate:

a) Vd
b) k
c) Concentration at 12 hours

### Solution 1

**a) Calculate Vd:**
$$V_d = \frac{Dose}{C_0} = \frac{1000\ mg}{25\ mg/L} = 40\ L$$

**b) Calculate k:**
$$k = \frac{0.693}{t_{1/2}} = \frac{0.693}{6\ h} = 0.116\ h^{-1}$$

**c) Concentration at 12 hours:**

12 hours = 2 half-lives

$$C_{12h} = C_0 \times (0.5)^2 = 25 \times 0.25 = 6.25\ mg/L$$

Or using exponential:
$$C_{12h} = 25 \times e^{-0.116 \times 12} = 6.25\ mg/L$$

### Problem 2

**Question:**

A drug has Cl = 50 mL/min. What infusion rate achieves Css = 10 mg/L?

### Solution 2

**Convert units:**
$$Cl = 50\ mL/min \times 60\ min/h \div 1000\ mL/L = 3\ L/h$$

**Calculate infusion rate:**
$$R_0 = C_{ss} \times Cl = 10\ mg/L \times 3\ L/h = 30\ mg/h$$

**Answer:** 30 mg/h

### Problem 3

**Question:**

If a drug has F = 0.5 and you need 200 mg to reach circulation, what oral dose should be given?

### Solution 3

**Formula:**
$$Dose_{oral} = \frac{Amount\ needed}{F}$$

**Calculation:**
$$Dose_{oral} = \frac{200\ mg}{0.5} = 400\ mg$$

**Answer:** 400 mg oral dose

**Verification:**
$$400\ mg \times 0.5 = 200\ mg\ reaches\ circulation\ \checkmark$$

:::

---

## Resources & Further Learning {.smaller}

::: {.columns}
::: {.column width="50%"}
**Boomer.org Chapters:**

📖 [Chapter 3: PK Fundamentals](https://www.boomer.org/c/p4/c03/c03.htm) - START HERE

📖 [Chapter 4: One-Compartment IV Bolus](https://www.boomer.org/c/p4/c04/c04.htm)

📖 [Chapter 5: Derived Parameters](https://www.boomer.org/c/p4/c05/c05.htm)

📖 [Chapter 6: IV Infusion](https://www.boomer.org/c/p4/c06/c06.htm)

📖 [Chapters 7-8: Oral Administration](https://www.boomer.org/c/p4/)

📖 [Full Syllabus](https://www.boomer.org/c/p4/syllabus.html)
:::

::: {.column width="50%"}
**Additional Resources:**

🔗 Interactive calculators on Boomer.org

🔗 [NCBI Pharmacokinetics](https://www.ncbi.nlm.nih.gov/books/NBK595006/)

🔗 [StatPearls: Drug Bioavailability](https://www.ncbi.nlm.nih.gov/books/NBK557852/)

🔗 [Deranged Physiology](https://derangedphysiology.com/)

**Reference Texts:**

📚 Rowland & Tozer: "Clinical PK & PD"

📚 Winter: "Basic Clinical PK"

📚 Shargel & Yu: "Applied Biopharm & PK"
:::
:::

---

## Questions? {.center}

::: {.r-fit-text}
Thank you!
:::

::: {.callout-tip icon=false}
## Contact & Feedback
- Office hours: [Time and location]
- Email: [Your email]
- Course website: [URL]
:::

::: {.fragment}
**Remember:** Pharmacokinetics is the foundation for rational drug therapy. Understanding these principles will make you a better clinician!
:::

---

## Bonus: Clinical Pearls {.smaller visibility="uncounted"}

::: {.callout-warning}
## Narrow Therapeutic Index Drugs
Small difference between therapeutic and toxic levels - monitor carefully!

- Warfarin
- Digoxin
- Lithium
- Phenytoin
- Theophylline
- Aminoglycosides
:::

::: {.callout-important}
## Red Flags

⚠️ **Saturable kinetics:** Phenytoin, aspirin, ethanol - can switch to zero-order

⚠️ **Highly protein bound:** Warfarin (99%), phenytoin (90%) - drug interactions affect free drug

⚠️ **Renally eliminated:** Adjust in renal impairment

⚠️ **Hepatically eliminated:** Adjust in liver disease
:::

---

## Appendix: Useful Conversions {.smaller visibility="uncounted"}

::: {.columns}
::: {.column width="50%"}
**Mathematical Constants:**

- ln(2) = 0.693
- e = 2.718

**Time:**

- 1 hour = 60 minutes
- 1 day = 24 hours = 1440 minutes

**Volume:**

- 1 L = 1000 mL
- 1 dL = 100 mL = 0.1 L
:::

::: {.column width="50%"}
**Concentration:**

- 1 mg/L = 1 mcg/mL = 1 μg/mL
- 1 g/dL = 10 g/L = 10,000 mg/L

**Body Water (70 kg adult):**

- Total body water: ~42 L (60%)
- Extracellular fluid: ~14 L (20%)
- Plasma volume: ~3 L (4%)
- Blood volume: ~5 L (7%)
:::
:::
